Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language

### 2022-08-06 revised_manuscript.tex
updated files

In our first set of hypotheses regarding referent tracking strategies,
we had the general expectation that signers would use more explicit
forms (mainly nominals) for referent introduction and less implicit
forms such as null marking, pronominal use, and classifiers. In
re-introduced contexts, we predicted that signers would use both
explicit and implicit markers depending on the accessibility of the
tracked referent.

For age of acquisition effects, based on what is proposed for child and
adult deaf signers with language deprivation in terms of their
morphosyntactic (Boudreault and Mayberry, 2006; Cormier et al., 2012;
Kayabaşı et al., 2020) and narrative development (Becker, 2009; Cormier
et al., 2013), we expected that late signers would be more
over-redundant in their referent tracking, akin to second language
learners (Bel et al., 2015; Frederiksen and Mayberry, 2019). That is, we
had the expectation that late signers would use more explicit markers
like nouns and fewer implicit markers like null marking even for highly
accessible referents (when the discourse status is maintained or even
re-introduced in same cases) than the control group (i.e., native
signers).

@@

However, the proportion of the observed discourse statuses (26\% for
introduction, 40\% for maintenance, 34\% for re-introduction across
acquisition groups) in this study differed from that of some other
studies. For a comparison, while the percentage of introduced referents
were similar in both research (26\% versus 24\%), the participants in
the present study brought back referents into the discourse more
frequently compared to the participants in Frederiksen and Mayberry
(2019) and Frederiksen and Mayberry (2016) which reported that
re-introduction contexts accounted for only 7\% of the referents in ASL.
On the other hand, maintenance percentages were lower in the present
study compared to Frederiksen and Mayberry (2016) (40\% versus 69\%).
Investigating the distribution of REs across discourse statuses in ASL,
a similar study by Czubek (2017) also reported contrasting numbers (17\%
for introduction, 64\% for maintenance, and 19\% for re-introduction).
We suggest that the discrepancy might be due to the nature of the
stimuli used in these research. In Frederiksen and Mayberry (2016), the
authors used 4-6-episode-long picture story with 2 animate characters
and 1 inanimate object. In contrast, the stimuli in the present study
consisted of 8-episode-long video clips on average with a mean number of
2 animate characters and 3 inanimate objects. Although the number of
animate characters and inanimate characters in Czubek (2017)'s stimuli
matched that of the present study, their narrative structure (a
6-episode-long picture story) was still shorter compared to our
narrative stimuli. We believe fewer possible referents and/or a simpler
narrative structure in these studies may have contributed to the lower
number of re-introduced contexts compared to the present study. Due to
the story length and composition in this study, the participants had to
track more recurring animate or inanimate referents for a longer period
of time. As a result, our signers engaged in referent re-introduction
more than what was reported in previous studies (Czubek, 2017;
Frederiksen and Mayberry, 2016).

As for the distribution of REs across different discourses statuses, we
observed that signers overwhelmingly preferred to use nominals over
other REs for the first mention of a referent and used null markers such
as CA, and agreement and plain verb for referents with high
accessibility (i.e., when the discourse status was maintained). This
finding parallels other accounts of referential accessibility suggested
for other sign languages {[}see Frederiksen and Mayberry (2016); Czubek
(2017); and Swabey (2002); for ASL, Morgan (2005); for BSL Bel et al.
(2015); and Barberà and Massó (2009) for LSC. Our results also indicate
that modified nouns (e.g., nouns that are preceded or followed by a
classifier or IX) appeared more in introduced contexts compared to
re-introduced contexts. This brings along the assumption that modified
nouns are used more to track referents with lower accessibility,
probably due to their more explicit structure.

Interestingly and to our surprise, the signers in this study used
pronominal IX very sporadically with only 40 observed instances that
were mainly used to re-introduce referents rather than maintain them.
This finding is comparable with what has been reported for ASL (Czubek,
2017; Frederiksen and Mayberry, 2016; Swabey, 2002) but not with LSC
(Bel et al., 2015), BSL (Morgan, 2005) and DGS (Perniss and Özyürek,
2015), in which the signers extensively made overt pronominal references
(up to 30\%) to maintain entities. Comparable to our findings, It has
been reported that pronouns do not frequently occur in the local context
and instead follows a topic shift and thus rank higher than classifiers
and null markers different genre (see Barberà and Massó, 2009 for LSC;
and Nuhbalaoglu, 2018 for TİD). Considering this, the claim that the IX
sign functions like demonstratives that are also used to track referents
in re-introduced contexts has not been borne out (Koulidobrova and
Lillo-Martin, 2016). Similar to these research, our findings indicate
that the pronominal IX in TİD was used more concurrently with nominal
constructions and pronominal IX was used very infrequently in isolation.
The observation that overt pronouns are rare in ASL narratives comes
from studies that include a simple picture retelling task with one
protagonist (e.g., Frederiksen and Mayberry, 2016) as well as a more
complex narrative design as in (Czubek, 2017). However, Czubek (2017)'s
study also incorporated different genres and reported more instances of
overt pronouns in personal narratives (compared to picture or video
retelling). In fact, this finding suggests that the use of pronouns
might be genre-dependent. Moreover, another production study on referent
tracking in ASL reported object preference for pronominals (Frederiksen
and Mayberry, 2017). Nonetheless, we only coded subjects for maintained
and re-introduced referents. We argue that all of these factors
(genre-dependency as well as object preference) might contribute to the
explanation of why overt pronouns in this data are dispreferred in favor
of other markers for highly accessible referents such as null marking
and classifiers.

A deeper look into the allocation of classifier constructions revealed
that WECLs were the most common type of classifiers for the maintenance
of a referent followed by BPCLs and ExtCLs. We observed a similar
alignment for re-introduced contexts. In contrast, a great number of the
overall ExtCLs were used to introduce a new referent into the discourse.
That is to say, whereas WECLs were the preferred classifier types to
signal highly accessible referents, ExtCLs constructions were mainly
used for referents which were not easily accessible. We suggest that the
less iconic and transparent nature of WECLs allowed signers to continue
a previous referent without being redundant. As pointed out by Czubek
(2017), signers prefer to rely on ExtCLs for the first mention of an
entity because these constructions are informative and explicit by
nature, and they provide visual and descriptive cues about an entity and
its location in space. Nevertheless, we only observed 3 instances of
ExtCLs for re-introduced contexts, replicating Frederiksen and Mayberry
(2019)'s findings for ASL.

As for the distribution of the null markers, approximately 90\% of the
plain verbs were used to maintain a previous referent, making these
constructions least implicit REs used for TİD\footnote{As in other sign
  languages, in TİD, the lexical meaning of signs has an effect on the
  morphological grouping. For instance, usually (di)transitive agreement
  verbs denote some sort of physical or metaphorical transfer of an
  object or abstract notion (Meir, 1999). As for plain verbs, both
  phonology of a sign and its semantics have an effect on morphology.
  For instance, \textsc{eat} in TİD is signed nearby the mouth and since
  it needs to remain in that phonological location, it is not suitable
  for agreement and thus it is regarded as a plain verb. Likewise the
  experiencer verb \textsc{love} is signed on the torso and as such it
  does not take part in agreement. These generalizations noted we should
  also be cautious in labelling a verb as categorially one type or the
  other. For instance, some intransitive plain verbs like \textsc{fall}
  or \textsc{jump} in TİD may display spatial person agreement with the
  single argument in the clause (Gökgöz and Sevgi, 2020). See Lourenço
  and Wilbur (2018) also for a similar discussion in Brazilian Sign
  Language.\}.}. Agreement verbs and CA were the following implicit
markers of reference, 0.77 and 0.72 of which were used for maintained
contexts, respectively. This contrasts with some of the research
(Czubek, 2017) in which agreement verbs ranked lowest on scale,
indicating that these constructions were more implicit compared to plain
verbs. However, there seems to be a common ground for the position of CA
markers which have been claimed to rank higher than other null markers
as in the present study (Barberà and Massó, 2009; Czubek, 2017;
Frederiksen and Mayberry, 2016). In Figure \ref{fig:scale}, we report
the distribution of REs in TİD as a function of referent accessibility
and compare our accessibility hierarchy with that of (Frederiksen and
Mayberry, 2016) for ASL. In the hierarchies, the left most REs (e.g.,
modified and bare nouns for TİD) are used when the accessibility is low.
If the referent is already active in the addressee's mind and its
accessibility is high, then the right most markers (e.g., agreement and
plain verbs for TİD) are selected.

\begin{figure}
\caption{Comparing the present study and Frederikse and Mayberry (2016) in terms of proposed accessibility hierarchies}
\label{fig:scale}
\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

@@

Given the paucity of research directed at the late signing group, we now
turn to the referent tracking strategies observed for second language
learners of a sign language. Our study replicated some of the findings
of the line of research that investigated deaf or hearing bimodal
bilingual signers (Bel et al., 2015; Frederiksen and Mayberry,
2019).\footnote{For similar results with heritage learners of sign language, see Reynolds (2018).}
The authors reported that more instances of null marking in the context
of referent introduction was attested for native signers than deaf or
hearing bimodal bilingual signers. Although using null marking to
introduce a referent into the discourse would normally result in
referential ambiguity, it is important to note that our stimuli
consisted of 10 interconnected narratives with 2 or more recurring
characters. It may be possible that native signers, having already
introduced the referent with an explicit nominal marker in a different
discourse context, found it felicitous to refer to the same entity with
more implicit markers in the following narratives even for introduced
contexts. Then, our results imply a difference in pragmatic sensitivity
to referential accessibility between native and late signers, which is
more robustly reflected in introduced contexts.

@@

The present study aimed to investigate the reference tracking strategies
of native and late deaf adult signers in TİD narratives. In line with
our expectations, we found that the distribution of REs (nominal,
pronominal, classifier, and null marking) followed from previous
accounts of saliency and referential accessibility. For introduced
contexts, signers produced more nominals and ExtCLs, and to maintain a
previous referent, hey mainly used null arguments (CA, and plain and
agreement verbs) and certain classifiers (e.g., BPCL and WECL) to track
highly accessible referents. We observed only very few instances of
pronouns (mostly produced by late signers), unlike our expectations. The
ones we observed were mostly reserved for re-introduced contexts along
with nominals. Based on our findings across the two groups, we proposed
a hierarchy of referent accessibility for TİD. As for the age of
acquisition effects, we observed limited over-redundancy for signers
with delayed language acquisition (i.e., late signers) who employed
fewer implicit forms of reference tracking compared to native signers.
Native signers were less likely to use a nominal for maintained contexts
and more likely to use null marking for introduced contexts than late
signers. We conclude that native signers might only partially adhere to
pragmatic norms and leverage accessibility better than late signers
since they use less form when the referent is more active in the mind.

>>>>

In our first set of hypotheses regarding referent tracking strategies,
we had the general expectation that signers would use more explicit
forms (mainly nominals) for referent introduction and less implicit
forms such as null marking, pronominal use, and classifiers for
maintenance. In re-introduced contexts, we predicted that signers would
use both explicit and implicit markers depending on the accessibility of
the tracked referent.

For age of acquisition effects, based on what is proposed for child and
adult deaf signers with language deprivation in terms of their
morphosyntactic (Boudreault and Mayberry, 2006; Cormier et al., 2012;
Kayabaşı et al., 2020) and narrative development (Becker, 2009; Cormier
et al., 2013), we expected that late signers would be more
over-redundant in their referent tracking, akin to second language
learners (Bel et al., 2015; Frederiksen and Mayberry, 2019). That is, we
had the expectation that late signers would use more explicit markers
like nouns and fewer implicit markers like null marking even for highly
accessible referents (when the discourse status is maintained or even
re-introduced in same cases) than native signers.

@@

However, the proportion of the observed discourse statuses (26\% for
introduction, 40\% for maintenance, 34\% for re-introduction across
acquisition groups) in this study differed from that of some other
studies. For a comparison, while the percentage of introduced referents
were similar in both research (26\% versus 24\%), the participants in
the present study brought back referents into the discourse more
frequently compared to the participants in Frederiksen and Mayberry
(2019) and Frederiksen and Mayberry (2016) which reported that
re-introduction contexts accounted for only 7\% of the referents in ASL.
On the other hand, maintenance percentages were lower in the present
study compared to Frederiksen and Mayberry (2016) (40\% versus 69\%).
Investigating the distribution of REs across discourse statuses in ASL,
a similar study by Czubek (2017) also reported contrasting numbers (17\%
for introduction, 64\% for maintenance, and 19\% for re-introduction).
We suggest that the discrepancy might be due to the nature of the
stimuli. In Frederiksen and Mayberry (2016), the authors used a
4-6-episode-long picture story with 2 animate characters and 1 inanimate
object. In contrast, the stimuli in the present study consisted of
8-episode-long video clips on average with a mean number of 2 animate
characters and 3 inanimate objects. Although the number of animate
characters and inanimate characters in Czubek (2017)'s stimuli matched
that of the present study, their narrative structure (a 6-episode-long
picture story) was still shorter compared to our narrative stimuli. We
believe fewer possible referents and/or a simpler narrative structure in
these studies may have contributed to the lower number of re-introduced
contexts compared to the present study. Due to the story length and
composition in this study, the participants had to track more recurring
animate or inanimate referents for a longer period of time. As a result,
our signers engaged in referent re-introduction more than what was
reported in previous studies (Czubek, 2017; Frederiksen and Mayberry,
2016).

As for the distribution of REs across different discourses statuses, we
observed that signers overwhelmingly preferred to use nominals over
other REs for the first mention of a referent and used null markers such
as CA, agreement and plain verb for referents with high accessibility
(i.e., when the discourse status was maintained). This finding parallels
other accounts of referential accessibility suggested for other sign
languages (see Frederiksen and Mayberry (2016); Czubek (2017); and
Swabey (2002); for ASL, Morgan (2005); for BSL Bel et al. (2015); and
Barberà and Massó (2009) for LSC). Our results also indicate that
modified nouns (e.g., nouns that are preceded or followed by a
classifier or IX) appeared more in introduced contexts compared to
re-introduced contexts. This brings along the assumption that modified
nouns are used more to track referents with lower accessibility,
probably due to their more explicit structure.

Interestingly and to our surprise, the signers in this study used
pronominal IX very sporadically with only 40 observed instances that
were mainly used to re-introduce referents rather than maintain them.
This finding is comparable with what has been reported for ASL (Czubek,
2017; Frederiksen and Mayberry, 2016; Swabey, 2002) but not with LSC
(Bel et al., 2015), BSL (Morgan, 2005) and DGS (Perniss and Özyürek,
2015), in which the signers extensively made overt pronominal references
(up to 30\%) to maintain entities. Comparable to our findings, it has
been reported that pronouns do not frequently occur in the local context
and instead follows a topic shift and thus rank higher than classifiers
and null markers (see Barberà and Massó, 2009 for LSC; and Nuhbalaoglu,
2018 for TİD). Considering this, the claim that the IX sign functions
like demonstratives that are also used to track referents in
re-introduced contexts has not been borne out (Koulidobrova and
Lillo-Martin, 2016). Similar to these research, our findings indicate
that the pronominal IX in TİD was used more concurrently with nominal
constructions and pronominal IX was used very infrequently in isolation.
The observation that overt pronouns are rare in ASL narratives comes
from studies that include a simple picture retelling task with one
protagonist (e.g., Frederiksen and Mayberry, 2016) as well as a more
complex narrative design as in (Czubek, 2017). However, Czubek (2017)
also incorporated different genres and reported more instances of overt
pronouns in personal narratives (compared to picture or video
retelling). In fact, this finding suggests that the use of pronouns
might be genre-dependent. Moreover, another production study on referent
tracking in ASL reported object preference for pronominals (Frederiksen
and Mayberry, 2017). Nonetheless, we only coded subjects for maintained
and re-introduced referents. We argue that all of these factors
(genre-dependency as well as object preference) might contribute to the
explanation of why overt pronouns in this data are dispreferred in favor
of other markers for highly accessible referents such as null marking
and classifiers.

A deeper look into the allocation of classifier constructions revealed
that WECLs were the most common type of classifiers for the maintenance
of a referent followed by BPCLs and ExtCLs. We observed a similar
distibution for re-introduced contexts. In contrast, a great number of
the overall ExtCLs were used to introduce a new referent into the
discourse. That is to say, whereas WECLs were the preferred classifier
types to signal highly accessible referents, ExtCL constructions were
mainly used for referents which were not easily accessible. We suggest
that the less iconic and transparent nature of WECLs allowed signers to
continue a previous referent without being redundant. As pointed out by
Czubek (2017), signers prefer to rely on ExtCLs for the first mention of
an entity because these constructions are informative and explicit by
nature, and they provide visual and descriptive cues about an entity and
its location in space. We only observed 3 instances of ExtCLs for
re-introduced contexts, replicating Frederiksen and Mayberry (2019)'s
findings for ASL.

As for the distribution of the null markers, approximately 90\% of the
plain verbs were used to maintain a previous referent, making these
constructions least implicit REs used for TİD\footnote{As in other sign
  languages, in TİD, the lexical meaning of signs has an effect on the
  morphological grouping. For instance, usually (di)transitive agreement
  verbs denote some sort of physical or metaphorical transfer of an
  object or abstract notion (Meir, 1999). As for plain verbs, both
  phonology of a sign and its semantics have an effect on morphology.
  For instance, \textsc{eat} in TİD is signed nearby the mouth and since
  it needs to remain in that phonological location, it is not suitable
  for agreement and thus it is regarded as a plain verb. Likewise the
  experiencer verb \textsc{love} is signed on the torso and as such it
  does not take part in agreement. These generalizations noted we should
  also be cautious in labelling a verb as categorially one type or the
  other. For instance, some intransitive plain verbs like \textsc{fall}
  or \textsc{jump} in TİD may display spatial person agreement with the
  single argument in the clause (Gökgöz and Sevgi, 2020). See Lourenço
  and Wilbur (2018) also for a similar discussion in Brazilian Sign
  Language.\}.}. Agreement verbs and CA were the following implicit
markers of reference, 0.77 and 0.72 of which were used for maintained
contexts, respectively. This contrasts with some of the research
(Czubek, 2017) in which agreement verbs ranked lowest on scale,
indicating that these constructions were more implicit compared to plain
verbs. However, there seems to be a common ground for the position of CA
markers which have been claimed to rank higher than other null markers
as in the present study (Barberà and Massó, 2009; Czubek, 2017;
Frederiksen and Mayberry, 2016). In Figure \ref{fig:scale}, we report
the distribution of REs in TİD as a function of referent accessibility
and compare our accessibility hierarchy with that of (Frederiksen and
Mayberry, 2016) for ASL. In the hierarchies, the leftmost REs (e.g.,
modified and bare nouns for TİD) are used when the accessibility is low.
If the referent is already active in the addressee's mind and its
accessibility is high, then the rightmost markers (e.g., agreement and
plain verbs for TİD) are selected.

\begin{figure}
\caption{Comparing the present study and Frederiksen and Mayberry (2016) in terms of proposed accessibility hierarchies}
\label{fig:scale}
\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

@@

Given the paucity of research directed at the late signing group, we now
turn to the referent tracking strategies observed for second language
learners of a sign language. Our study replicated some of the findings
of the line of research that investigated second language learners of a
sign language (Bel et al., 2015; Frederiksen and Mayberry, 2019). The
authors reported that more instances of null marking in the context of
referent introduction was attested for native signers than second
language learners. Although using null marking to introduce a referent
into the discourse would normally result in referential ambiguity, it is
important to note that our stimuli consisted of 10 interconnected
narratives with 2 or more recurring characters. It may be possible that
native signers, having already introduced the referent with an explicit
nominal marker in a different discourse context, found it felicitous to
refer to the same entity with more implicit markers in the following
narratives even for introduced contexts. Then, our results imply a
difference in pragmatic sensitivity to referential accessibility between
native and late signers, which is more robustly reflected in introduced
contexts. To test this implication, one could use the same number and
length of stimuli from different cartoons with different characters and
see if the tendency to introduce a referent without an overt nominal for
the first time is dependent on the task or not.

@@

The present study aimed to investigate the reference tracking strategies
in TİD narratives and whether native and late deaf TİD signers differ
from each other in these strategies. In line with our expectations, we
found that the distribution of REs (nominal, pronominal, classifier, and
null marking) followed from previous accounts of saliency and
referential accessibility. For introduced contexts, signers produced
more nominals and ExtCLs, and to maintain a previous referent, they
mainly used null arguments (in clauses with CA, agreement and plain
verbs) and certain classifiers (e.g., BPCL and WECL) to track highly
accessible referents. We observed only very few instances of pronouns
(mostly produced by late signers), unlike our expectations. The ones we
observed were mostly reserved for re-introduced contexts along with
nominals. Based on our findings across the two groups, we proposed a
hierarchy of referent accessibility for TİD. As for the age of
acquisition effects, we observed limited over-redundancy for signers
with delayed language acquisition (i.e., late signers) who employed
fewer implicit forms of reference tracking compared to native signers.
Native signers were less likely to use a nominal for maintained contexts
and more likely to use null marking for introduced contexts than late
signers. Following Swabey (2002)'s intituition, we conclude that native
signers must adhere to pragmatic norms and leverage accessibility better
than late signers only to a certain extent (given our finding of limited
overexplicitness for late signers) since they use less form when the
referent is more active in the mind.
